---
title: "Background on ACS NY data with PUMS"
output: github_document
---
  
### Econ B2000, MA Econometrics
### Kevin R Foster, the Colin Powell School at the City College of New York, CUNY

### A different dataset...
  
Now we will pivot and use a different dataset, ACS data on wage, education (and lots of other things). You can get this data from the [course page](https://kfoster.ccny.cuny.edu/classes/fall2022/).

We will work on an extract with only people who live in the state of New York. It includes quite granular geographic information, getting down to a person's neighborhood (referred to as PUMA). Well, that's not quite true. In NYC PUMA means a relatively small area although in upstate, where the population density is so much lower, a PUMA could be quite large in order to collect together enough people. In NYC, PUMA is a neighborhood.

The PUMA codes are cryptic. You have to go to the codebook (or, in this case, the file PUMA_levels.csv or acs2017_codebook.txt from the zip file) to find out that 3801 codes for Washington Heights/Inwood, 3802 is Hamilton Heights/Manhattanville/West Harlem, etc. The program will happily calculate the average value for PUMA (type in *mean(PUMA)* and see for yourself!) but this is a meaningless value -- wtf is the average neighborhood code value!? A PUMA value here is like a zip code, it's a number but you wouldn't want to do number-type operations such as taking a mean. (Although R would let you do that.)

If you want to select just people living in a particular neighborhood then you'd have to look at the list below. 

PUMA | Neighborhood
---- | ------------
3701 | NYC-Bronx CD 8--Riverdale, Fieldston & Kingsbridge
3702 | NYC-Bronx CD 12--Wakefield, Williamsbridge & Woodlawn
3703 | NYC-Bronx CD 10--Co-op City, Pelham Bay & Schuylerville
3704 | NYC-Bronx CD 11--Pelham Parkway, Morris Park & Laconia
3705 | NYC-Bronx CD 3 & 6--Belmont, Crotona Park East & East Tremont
3706 | NYC-Bronx CD 7--Bedford Park, Fordham North & Norwood
3707 | NYC-Bronx CD 5--Morris Heights, Fordham South & Mount Hope
3708 | NYC-Bronx CD 4--Concourse, Highbridge & Mount Eden
3709 | NYC-Bronx CD 9--Castle Hill, Clason Point & Parkchester
3710 | NYC-Bronx CD 1 & 2--Hunts Point, Longwood & Melrose
3801 | NYC-Manhattan CD 12--Washington Heights, Inwood & Marble Hill
3802 | NYC-Manhattan CD 9--Hamilton Heights, Manhattanville & West Harlem
3803 | NYC-Manhattan CD 10--Central Harlem
3804 | NYC-Manhattan CD 11--East Harlem
3805 | NYC-Manhattan CD 8--Upper East Side
3806 | NYC-Manhattan CD 7--Upper West Side & West Side
3807 | NYC-Manhattan CD 4 & 5--Chelsea, Clinton & Midtown Business District
3808 | NYC-Manhattan CD 6--Murray Hill, Gramercy & Stuyvesant Town
3809 | NYC-Manhattan CD 3--Chinatown & Lower East Side
3810 | NYC-Manhattan CD 1 & 2--Battery Park City, Greenwich Village & Soho
3901 | NYC-Staten Island CD 3--Tottenville, Great Kills & Annadale
3902 | NYC-Staten Island CD 2--New Springville & South Beach
3903 | NYC-Staten Island CD 1--Port Richmond, Stapleton & Mariner's Harbor
4001 | NYC-Brooklyn CD 1--Greenpoint & Williamsburg
4002 | NYC-Brooklyn CD 4—Bushwick
4003 | NYC-Brooklyn CD 3--Bedford-Stuyvesant
4004 | NYC-Brooklyn CD 2--Brooklyn Heights & Fort Greene
4005 | NYC-Brooklyn CD 6--Park Slope, Carroll Gardens & Red Hook
4006 | NYC-Brooklyn CD 8--Crown Heights North & Prospect Heights
4007 | NYC-Brooklyn CD 16--Brownsville & Ocean Hill
4008 | NYC-Brooklyn CD 5--East New York & Starrett City
4009 | NYC-Brooklyn CD 18--Canarsie & Flatlands
4010 | NYC-Brooklyn CD 17--East Flatbush, Farragut & Rugby
4011 | NYC-Brooklyn CD 9--Crown Heights South, Prospect Lefferts & Wingate
4012 | NYC-Brooklyn CD 7--Sunset Park & Windsor Terrace
4013 | NYC-Brooklyn CD 10--Bay Ridge & Dyker Heights
4014 | NYC-Brooklyn CD 12--Borough Park, Kensington & Ocean Parkway
4015 | NYC-Brooklyn CD 14--Flatbush & Midwood
4016 | NYC-Brooklyn CD 15--Sheepshead Bay, Gerritsen Beach & Homecrest
4017 | NYC-Brooklyn CD 11--Bensonhurst & Bath Beach
4018 | NYC-Brooklyn CD 13--Brighton Beach & Coney Island
4101 | NYC-Queens CD 1--Astoria & Long Island City
4102 | NYC-Queens CD 3--Jackson Heights & North Corona
4103 | NYC-Queens CD 7--Flushing, Murray Hill & Whitestone
4104 | NYC-Queens CD 11--Bayside, Douglaston & Little Neck
4105 | NYC-Queens CD 13--Queens Village, Cambria Heights & Rosedale
4106 | NYC-Queens CD 8--Briarwood, Fresh Meadows & Hillcrest
4107 | NYC-Queens CD 4--Elmhurst & South Corona
4108 | NYC-Queens CD 6--Forest Hills & Rego Park
4109 | NYC-Queens CD 2--Sunnyside & Woodside
4110 | NYC-Queens CD 5--Ridgewood, Glendale & Middle Village
4111 | NYC-Queens CD 9--Richmond Hill & Woodhaven
4112 | NYC-Queens CD 12--Jamaica, Hollis & St. Albans
4113 | NYC-Queens CD 10--Howard Beach & Ozone Park
4114 | NYC-Queens CD 14--Far Rockaway, Breezy Point & Broad Channel

So you want to tell R to treat PUMA as a factor.
```{r}
load("acs2017_ny_data.RData")
attach(acs2017_ny)
PUMA <- as.factor(PUMA)
```
And you might want to do that for some of the others too.

I will leave you to worry over the recoding of the other variables, because it's good for the soul. I will show you 2 ways -- the quick and dirty way, and the fancy correct way.

First the quick and dirty way.
```{r}
female <- as.factor(female)
print(levels(female))
levels(female) <- c("male","female")
```
Well, ways,
```{r}
educ_indx <- factor((educ_nohs + 2*educ_hs + 3*educ_somecoll + 4*educ_college + 5*educ_advdeg), levels=c(1,2,3,4,5),labels = c("No HS","HS","SmColl","Bach","Adv"))
```

(If you can figure out how that bit of code works, that would be good)

These just type in the levels. But for things like PUMA, it could be a long list and might not even match every one. To do it better, we need help from an R package.

```{r message=FALSE}
library(tidyverse)
library(plyr)
levels_n <- read.csv("PUMA_levels.csv")
levels_orig <- levels(PUMA) 
levels_new <- join(data.frame(levels_orig),data.frame(levels_n))
levels(PUMA) <- levels_new$New_Level
```

Those commands read in a little csv file that I had made, with the PUMA codes, then matches the old codes with the new complete text. Note that I'm lazy so codes in NY state outside of NYC are coded NA.

R will do the summary differently when it knows the variable is a factor,
```{r}
summary(PUMA)
summary(educ_indx)
```

To find mean and standard deviation by neighborhood, you could use something like this,

```{r}
ddply(acs2017_ny, .(PUMA), summarize, mean = round(mean(AGE), 2), sd = round(sd(AGE), 2), n_obsv = length(PUMA))
```
Although tapply would also work fine.

Here's the 90th and 10th percentiles of wages by neighborhood,

```{r}
dat_use1 <- subset(acs2017_ny,((INCWAGE > 0) & in_NYC))
ddply(dat_use1, .(PUMA), summarize, inc90 = quantile(INCWAGE,probs = 0.9), inc10 = quantile(INCWAGE,probs = 0.1), n_obs = length(INCWAGE))
```

You could also use table (or crosstabs) for factors with fewer items,

```{r}
table(educ_indx,female)
xtabs(~educ_indx + female)
```
Want proportions instead of counts?
```{r}
prop.table(table(educ_indx,female))
```

Try it and see what happens if you use table with PUMA...

This data includes not just whether a person has a college degree but also what field was the degree in: Economics or Psychology, for instance. Look over the codebook about DEGFIELD and DEGFIELDD (that second D means more detail) to see the codes. Maybe look at 10th and 90th percentiles by degree field?
  
```{r}
detach()
```
